Refinement for open automata

### 2021-06-16 2021-Refinement/report.tex
Many corrections, still no TRVE conclusion

\section{Introduction}
The open automata model was originally designed to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification \cite{henrio:01252323}.
This model is based around a hierarchical composition of generalised automata.
Verification is the process of checking properties on a description in a specific model (for instance an automaton).
This properties can be for instance being in some type or being deadlock free (a deadlock is an unintended state where no further action is possible).
Open pNets with the open automaton semantics and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to model and verify some properties of applications \cite{qin:01823507, ameurboulifa:01526055} based on BIP (Behaviour-Interaction-Priority) \cite{basu:hal-00722395} and GCM (Grid Component Model) \cite{baude:inria-00323919}.
However equivalence relations are not always sufficient to perform verification, refinement relations are also used.
A refinement is the verification that an implementation satisfies a specification, whereas an equivalence would check that their behaviours are the same.
For instance a usecase on open pNet where equivalence is not sufficient can be found in an article about algorithms fo weak bisimulation on open automata \cite{wang:03126313}.

The contributions of this report are:
\begin{enumerate}
\item The formalisation of composition for open automata without requiring open pNets.
\item The translation of some properties of refinement relations from LTS to open automata.
\item The characterisation of a composition that doesn't introduce deadlocks.
\item The introduciton of several refinement relations for open automata that behave nicely with composition.
\end{enumerate}
% TODO: Related work: thesis

We begin with Section \ref{sec:notations} by giving some notations used throughout this paper.
Then there are several sections dedicated to define all the interesting objects and properties:
In Section \ref{sec:def} we give a clean and up to date definition of open automata.
In Section \ref{sec:comp} we define their composition without requiring the use of open pNets \Ludo{}{does this mean that previous definitions were using open pnets? this might be clarified in contributions but when you redo a definition you must highlight why a new definition was necessary}\Quentin{}{Yes, composition of open automata was composition of the open pNets generating them, the necessity is as stated, being independant from open pNets}.
In Section \ref{sec:proofelts} we adapt to open automata several standard properties from refinement relations on LTS.
After that Section \ref{sec:prelref} is dedicated to progressively build a refinement relation and explain the elements of this construction.
Section \ref{sec:refinement} introduces and analyses the refinement relation on open automata.
Finally Section \ref{sec:ccl} recapitulates the contribution and presents the work left.

@@

Families of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter can only be used if there is a \Ludo{generating expression}{c'est vrai pour les exemples mais pas pour les defs generiques (OA, bisim, ...): reformuler}\Quentin{}{Je pense que c'est aussi vrai pour les defs génériques, on définit une famille à l'aide d'une expression génératrice, par exemple \(\mpar{\beta_j}_{j \in J}\) est en réalité \(j \mapsto \beta_j\); une fonction est une expression génératrice}; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
However \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no generating expression and is represented here with the finite version of first notation.

The disjoint union on set is noted \(\uplus\).
There are several ways of ensuring a union is disjoint, we will indifferently either suppose sets are disjoint or rename conflicting object (useful for variables).
The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) has the following signature \(\varphi \uplus \psi: I \uplus J \to X \cup Y\).

@@

Notations are parameterised by a term algebra \(\terms\).
As any term algebra it has constant symbols with arity, variables, and a typing mechanism to distinguish well-formed and ill-formed terms.
The action algebra \(\actions\) is another term algebra.
It \Ludo{can be}{not much informative}\Quentin{}{Intended use, it is not mandatory} a subset of the terms.
The formulas \(\formulas\) are at least the first order formulas over \(\terms\) and \(\actions\).
\end{defi}
The term algebras are arbitrary.
The formulas contain at least first order logic terms with an equality relation.
This equality relation is not necessarily a syntactic equality (\(2 + 2 \neq 4\) with a syntactic equality but it is more expressive to consider that the two terms are equal).

@@

The interpretation of terms is supposed to be decidable.
The satisfiability of formulas might not be decidable nor complete nor consistent,
but we will consider in the following that they are; this can be achieved by restricting the term algebra to a decidable subset.
For instance a formula with quantifiers on variables might not be provable even if it is true for all values of these variables.
In practice the formulas will be given to a SMT solver and we cannot always make sure they have all the previous properties but, as soon as the satisfiability is consistent our theory is consistent too.
\(\vdash\) can hence be interpreted as an indicator of what is given to the SMT; it separates the external logic and the logic on \(\formulas\).

Values will be used for keeping a variable state, and then injected in terms for substitution.
This is correct when \(\values \subseteq \rterms\) and we suppose it is the case.
Otherwise it doesn't invalidates theorems because it can stand as a shorthand for substitution with any term which has the kept value as an interpretation.
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
\begin{noti}[Notations for separating external logic and logic on \(\formulas\)]
\defitem The satisfiability of a formula \(f \in \formulas\) under some valuation \(\sigma: V \to \values\) is noted:
\[ \sigma \vdash f \defnotation \vdash \exists \fvars{f\psubst{\sigma}}, f\psubst{\sigma} \]
\defitem The satisfiability of a formula \(f \in \formulas\) with some variable set \(V\) as context is noted:
\[ V \vdash f \defnotation \vdash \forall V, \exists\mpar{\fvars{f} \setminus V}, f \]
\defitem The precedence of \(\vdash\) is the lowest on the right side and higher than \(\uplus\) on the left side:
\[ \forall a \, b, a \uplus b \vdash x \wedge y \implies \exists z, P\mpar{x,z} \text{ is the same as } \forall a \, b, \bigg((a \uplus b) \vdash \Big((x \wedge y) \implies \exists z, P\mpar{x,z}\Big)\bigg) \]
\end{noti}
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A open automaton is a tuple \(\OAg\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.
A pair of a state and a valuation is called a configuration.

@@

The initial valuation is a partial valuation of the variables.
If an undefined variable is set before being read, it behaves like an initially set variable.
If an undefined variable is not set before being read then its value may be any fixed value.
When we will introduce refinements in Section \ref{sec:prelref}, setting an undefined variable will be considered a valid refinement, for instance a \(5\) bits register is a particular \(n\) bits register.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(\beta_j \in \actions\) the actions of the holes, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variable assignments.
\end{defi}
An open transition can have many unbound variables.
Actually an effective transition of the automaton is any well-formed substitution of the unbound variables of the transition minus the automaton variables.

The intuition of an open automaton is a partially defined LTS with variables, guards on transitions, and parametrised actions. The LTS is partial in the sense that one cn plug an LTS in the hole and the surrounding LTS will react to the actions emitted by the sub-LTS put in the hole.
Initially the automaton is in the initial state with an extension on all the variables of the initial valuation.
In any configuration, it can perform effective transitions which source state is the current state and which guard is satisfiable in the current valuation if the holes emit the indicated actions.
An effective transition is a transition where every variable that is not an automaton variable is instantiated with a value.
By performing the effective transition, the automaton emits the indicated action and updates its configuration according to the target state and variable assignments.

@@

\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator implementation with open automata, on the left state oriented, on the right data oriented}
\label{fig:enable}
\end{figure}
The automaton on the left side of the figure is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{\forall y, x \neq \delta\mpar{y}}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
Note that the transition in the middle could have been expressed as \(\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{x}}\).
It would have avoided many effective transitions with trivially false guards like the one where \(x \mapsto 1, y \mapsto 2\), which has the guard \(1 = 2\). \Ludo{}{I believe that with the usual definition of the semantics of open automata it is the same ... please check: you do all potential transitions with the conditions that x is the same in both. In any case as the matching is not linear, the simplification is not obvious}\Quentin{}{Same behaviour as here with the standard definition: RR-9177, sec 5.2}

@@

We use the standard convention that uses \(\tau\) as a non-observable transition, never synchronised with other actions and passed unmodified.
This allows the synchronisation of the two holes to be hidden to the exterior by sending a \(\tau\).
However here \(\tau\) is not always allowed from the holes, for instance in the state \(L\) the hole \(r\) cannot emit it.
These transitions have been omitted for the sake of simplifying the first example of open automata.
\end{exi}
Finally we can define some utilitary functions:
\begin{defi}[Guard, Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT{r}\) are called the out-transitions of \(r\).
\(\fIT{r}\) are called the in-transitions of \(r\).
The local variables of a transition are all variables appearing in that transition except the global variables of the automaton.
\begin{align*}
	\fOT{r} & \defnotation \mset{\OTg \in T}{s = r} &
	\fIT{r} & \defnotation \mset{\OTg \in T}{s' = r} \\
\end{align*}
\vspace{-1cm}
\begin{gather*}
	\fguard{\OTg} \defnotation g \\
	\fvars{\OTg} \defnotation \mpar{\fvars{\alpha} \cup \fvars{g} \cup \bigcup_{j \in J'} \fvars{\beta_j} \cup \bigcup_{v \in V} \fvars{\psi\mpar{v}}} \setminus V
\end{gather*}
\end{defi}
The goal of the extractor \(\fvars{t}\) when \(t\) is a transition is to get variables that are only present in that transition.
Also transition variables sets are disjoint from one transition to another.
Other variables like automaton variables are already known at this point so there is no use in getting their name again.
Also the use of this extractor benefits from this exclusion: otherwise there would always be \(\setminus V\) following it to prevent variable shadowing.

\Ludo{}{is it really variable shadowing?}\Quentin{}{I don't know if the term is the same in math}

\Ludo{}{you do not necessarily need to explain everything if this is what you need, but you miss an intuition of what this set is}\Quentin{}{I don't know what else I could say to give the intuition, all elements are here, in my opinion, the only thing left to build the intuition is an example}

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{\mbrc{v \mapsto l}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = L \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = L \wedge x = y}{\mbrc{v \gets R}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = R}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is \Ludo{FH-bisimilar}{makes no sense: FH bisimlar is still undefined, neither formally nor informally} to the previous one \cite{henrio:01299562}.\Quentin{}{I can remove this sentence, it adds nothing}
The variables that are said to be local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
When taking effective transitions, its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = L \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{L = L \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto L\)).

@@

Note that in the previous paragraph there are two relations:
The first is between states of two specific automata (\(R \subseteq S_1 \times S_2\)), the second is between automata (\(A_1 \leq A_2\)).
Generally the second relation is implicit because its definition is always \(\exists R \subseteq S_1 \times S_2, P\mpar{R}\) with \(P\) a property on the relation on states, so only talking about \(R\) by defining \(P\) is sufficient.\Ludo{}{je comprends pas cette phrase}\Quentin{}{Est-ce mieux?}
However later on in this report we will define more complex relations on automata so we clarify the distinction now.

@@

Two states \(s_1 \in S_1, s_2 \in S_2\) with their respective valuations \(\sigma_1: V_1 \to \values, \sigma_2: V_2 \to \values\) are related iff \(\sigma_1 \uplus \sigma_2 \vdash R\mpar{s_1, s_2}\).
\end{defi}
In this section we give properties on refinement relations, both on the relations on configurations and on the relation on automata.
When relating two automata \(A_1 \mathrel{\mathcal{R}} A_2\) or their configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\), \(A_1\) will be called the implementation and \(A_2\) the specification.

For the relation on automata, the important properties we consider are:
\begin{defi} A relation on open automata \(\leq\) is
\defitem \textbf{reflexive} iff \(\forall a, a \leq a\);
\defitem \textbf{transitive} iff \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\defitem \textbf{a preorder} iff it is reflexive and transitive;
\defitem \textbf{correct w.r.t.\@ composition} iff \(\forall a\, b, a\subst{b}{j} \leq a\);
\defitem \textbf{complete w.r.t.\@ composition} iff \(\forall a\, b, a \leq b \implies \exists c, a \cong b\subst{c}{j}\);
\defitem \textbf{context refining for composition} iff \(\forall a\, b\, c, a \leq b \implies a\subst{c}{j} \leq b\subst{c}{j}\);
\defitem \textbf{congruent for composition} iff \(\forall a\, b\, c, a \leq b \implies c\subst{a}{j} \leq c\subst{b}{j}\);
\defitem \textbf{compatible with composition} iff \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\subst{a}{j} \leq d\subst{b}{j}\);
\defitem \textbf{compatible with FH-bisimulation} iff \(\forall a\, b\, c, d, a \cong b \wedge c \cong d \wedge a \leq c \implies b \leq d\).
\end{defi}
Reflexivity, transitivity and preorder are classical properties on relations.
Correctness w.r.t.\@ composition means that every composition is considered a refinement.
This property captures the expected behaviour of a refinement relation on a compositional structure like the open automata.
Completeness w.r.t.\@ composition means that a refinement corresponds to the left automaton being equivalent to some composition of the right automaton.
This property is useful if we want to characterise the kind of operations that can generate any refined automaton starting from the specification.
We won't have it because extending the domain of definition of the initial valuation will be a refinement which can't be expressed with composition.
Yet a ``completeness w.r.t.\@ some set of operations" is still a strong and interesting property.
Context refinement is the refinement being compatible with composition of the same automaton.
Congruence is the refinement being compatible with being composed in the same automaton.
Compatibility with composition is the conjunction of the two latter (assuming the relation is a preorder).
This property is required on compositional structures to be of any use in verification, proofs and for model checking.
Finally compatibility with FH-bisimulation is the refinement not being able to distinguish two FH-bisimilar automata.
Since FH-bisimulation is the pre-existing equivalence relation, we should not introduce an incompatible proof strategy.

>>>>

\section{Introduction}
The open automata model was originally designed to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification of concurrent systems \cite{henrio:01252323}.
Verification is the process of checking properties on a description in a specific model (for instance classical automata).
A particular kind of verification are refinement relations.

Refinement relations verifies that and implementation satisfies some properties with respect to a specification.
The open pNets model is based around a hierarchical composition of generalised automata called pLTS.
A properties a refinement can check on them is the non introduction of deadlocks.
A deadlock being an unintended state where no further action is possible.

The composition of open automata is inherited from open pNets as the open automaton generated from the composition on open pNets.
This means that composition is only defined on open pNets and not directly on open automata. \Quentin{}{Clunky transition}
Open pNets with the open automaton semantics and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to model and verify some properties of applications \cite{qin:01823507, ameurboulifa:01526055} based on BIP (Behaviour-Interaction-Priority) \cite{basu:hal-00722395} and GCM (Grid Component Model) \cite{baude:inria-00323919}.
However only an equivalence relation was used.
For instance a usecase on open pNet where equivalence is not sufficient can be found in an article about algorithms fo weak bisimulation on open automata \cite{wang:03126313}.

The contributions of this report are:
\begin{enumerate}
\item The formalisation of composition for open automata without requiring open pNets.
\item The translation of some properties of refinement relations from LTS to open automata.
\item The characterisation of a composition that doesn't introduce deadlocks.
\item The introduciton of several refinement relations for open automata that behave nicely with composition.
\end{enumerate}

This report is organised as follows:
We begin with Section \ref{sec:notations} by giving some notations used throughout this paper.
Then there are several sections dedicated to define all the interesting objects and properties:
In Section \ref{sec:def} we give a clean and up to date definition of open automata.
In Section \ref{sec:comp} we define their composition without requiring the use of open pNets.
In Section \ref{sec:proofelts} we adapt to open automata several standard properties from refinement relations on LTS.
After that Section \ref{sec:prelref} is dedicated to progressively build a refinement relation and explain the elements of this construction.
Section \ref{sec:refinement} introduces and analyses the refinement relation on open automata.
Finally Section \ref{sec:ccl} recapitulates the contribution and presents the work left.

@@

Families of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
For instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
They will be used depending on what is more convenient.
For instance \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no simple generating expression and is better represented with the finite version of first notation.

The disjoint union on set is noted \(\uplus\).
There are several ways of ensuring a union is disjoint, we will indifferently either suppose sets are disjoint or rename conflicting object (useful for variables).
The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is noted \(\varphi \uplus \psi\) and has the following signature \(I \uplus J \to X \cup Y\).

@@

Notations are parameterised by a term algebra \(\terms\).
As any term algebra it has constant symbols with arity, variables, and a typing mechanism to distinguish well-formed and ill-formed terms.
The action algebra \(\actions\) is another term algebra.
The formulas \(\formulas\) are at least the first order formulas over \(\terms\) and \(\actions\).
\end{defi}
The term algebras are arbitrary.
The formulas contain at least first order logic terms with an equality relation.
This equality relation is not necessarily a syntactic equality (\(2 + 2 \neq 4\) with a syntactic equality but it is more expressive to consider that the two terms are equal).

@@

The interpretation of terms is supposed to be decidable.
The satisfiability of formulas might not be decidable nor complete nor consistent,
but we will consider in the following that they are; this can be achieved by restricting the term algebra to a decidable subset.
For instance a formula with quantifiers on variables might not be provable even if it is true for all values of these variables.
In practice the formulas will be given to a SMT solver and we cannot always make sure they have all the previous properties but, at least any property found satsfiable by the SMT is ensured to be true.
\(\vdash\) can hence be interpreted as an indicator of what is given to the SMT; it separates the external logic and the logic on \(\formulas\).

Values will be used for keeping a variable state, and then injected in terms for substitution.
This is correct when \(\values \subseteq \rterms\) and we suppose it is the case.
Otherwise this doesn't invalidates theorems because substitution by a value can stand as a shorthand for substitution with any term which has the value as an interpretation.
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
\begin{noti}[Notations for separating external logic and logic on \(\formulas\)]
\defitem The satisfiability of a formula \(f \in \formulas\) under some valuation \(\sigma: V \to \values\) is noted:
\[ \sigma \vdash f \defnotation \vdash \exists \fvars{f\psubst{\sigma}}, f\psubst{\sigma} \]
\defitem The satisfiability of a formula \(f \in \formulas\) with some variable set \(V\) as context is noted:
\[ V \vdash f \defnotation \vdash \forall V, \exists\mpar{\fvars{f} \setminus V}, f \]
\defitem The precedence of \(\vdash\) is the lowest on the right side and higher than \(\uplus\) on the left side:
\[ \forall a \, b, a \uplus b \vdash x \wedge y \implies \exists z, P\mpar{x,z} \text{ is the same as } \forall a \, b, \bigg((a \uplus b) \vdash \Big((x \wedge y) \implies \exists z, P\mpar{x,z}\Big)\bigg) \]
\end{noti}
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A open automaton is a tuple \(\OAg\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.
A pair of a state and a valuation is called a configuration.

@@

The initial valuation is a partial valuation of the variables.
If an undefined variable is set before being read, it behaves like an initially set variable.
If an undefined variable is not set before being read then its value may be any fixed value.
When we will introduce refinements in Section \ref{sec:prelref}, setting an undefined variable will be considered a valid refinement, for instance a \(5\) bits register is a particular \(n\) bits register.
\begin{defi}[Open transition, Instanciated transition]
An open transition is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(\beta_j \in \actions\) the actions of the holes, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variable assignments.
An open transition can have many unbound variables.

An instanciated transition of the automaton is transition produced by any well-formed substitution of the unbound variables of the transition minus the automaton variables.
\end{defi}
The intuition of an open automaton is a partially defined LTS with variables, guards on transitions, and parametrised actions. The LTS is partial in the sense that one cn plug an LTS in the hole and the surrounding LTS will react to the actions emitted by the sub-LTS put in the hole.
Initially the automaton is in the initial state with an extension on all the variables of the initial valuation.
In any configuration, it can perform instanciated transitions which source state is the current state and which guard is satisfiable in the current valuation if the holes emit the indicated actions.
An instanciated transition is a transition where every variable that is not an automaton variable is instantiated with a value.
By performing the instanciated transition, the automaton emits the indicated action and updates its configuration according to the target state and variable assignments.

@@

\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator implementation with open automata, on the left state oriented, on the right data oriented}
\label{fig:enable}
\end{figure}
The automaton on the left side of the figure is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{\forall y, x \neq \delta\mpar{y}}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
Note that the transition in the middle could have been expressed as \(\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{x}}\).
It would have avoided many instanciated transitions with trivially false guards like the one where \(x \mapsto 1, y \mapsto 2\), which has the guard \(1 = 2\).

@@

We use the standard convention that uses \(\tau\) as a non-observable transition, never synchronised with other actions and passed unmodified.
This allows the synchronisation of the two holes to be hidden to the exterior by sending a \(\tau\).
However here \(\tau\) is not always allowed from the holes, for instance in the state \(L\) the hole \(r\) cannot emit it.
These transitions have been omitted for the sake of simplifying the first example of open automata.
\end{exi}
Finally we can define some utilitary functions:
\begin{defi}[Guard, Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT[T]{r} \in T\) are called the out-transitions of the state \(r\).
\(\fIT[T]{r} \in T\) are called the in-transitions of the state \(r\).
When the transition set is clear from the context, it will be omitted.
The local variables of a transition \(\fvars{t}\) are all variables appearing in transition \(t\) except the global variables of the automaton.
\begin{align*}
	\fOT[T]{r} & \defnotation \mset{\OTg \in T}{s = r} &
	\fIT[T]{r} & \defnotation \mset{\OTg \in T}{s' = r} \\
\end{align*}
\vspace{-1cm}
\begin{gather*}
	\fguard{\OTg} \defnotation g \\
	\fvars{\OTg} \defnotation \mpar{\fvars{\alpha} \cup \fvars{g} \cup \bigcup_{j \in J'} \fvars{\beta_j} \cup \bigcup_{v \in V} \fvars{\psi\mpar{v}}} \setminus V
\end{gather*}
\end{defi}
The goal of the extractor \(\fvars{t}\) when \(t\) is a transition is to get variables that are only present in that transition.
Also transition variables sets are disjoint from one transition to another.
if we need to quantify on other variables then the quantification will be done separately.
\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{\mbrc{v \mapsto l}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = L \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = L \wedge x = y}{\mbrc{v \gets R}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = R}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator.\footnote{It was proved FH-bisimilar to the previous one \cite{henrio:01299562}}
The variables that are said to be local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
When taking instanciated transitions, its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = L \wedge \tau \neq \delta}{\mbrc{}}} is an instanciated transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{L = L \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto L\)).

@@

Note that in the previous paragraph there are two relations:
The first is between states of two specific automata (\(R \subseteq S_1 \times S_2\)), the second is between automata (\(A_1 \leq A_2\)).
Generally the second relation is implicit because its definition is always \(\exists R \subseteq S_1 \times S_2, P\mpar{R}\) with \(P\) a property on the relation on states, so only talking about \(R\) by defining \(P\) is sufficient.
However later on in this report we will define more complex relations on automata so we clarify the distinction now.

@@

Two states \(s_1 \in S_1, s_2 \in S_2\) with their respective valuations \(\sigma_1: V_1 \to \values, \sigma_2: V_2 \to \values\) are related iff \(\sigma_1 \uplus \sigma_2 \vdash R\mpar{s_1, s_2}\).
\end{defi}
When relating two automata \(A_1 \mathrel{\mathcal{R}} A_2\) or their configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\), \(A_1\) will be called the implementation and \(A_2\) the specification.

In this section we first give expected properties for a refinement relation on open automata, then expected properties for a relation on configurations.
Finally a conflict between the properties is underlined and solved.

\subsection{Properties of the relation on automata}
The important properties we consider for a refinement relation on open automata are:
\begin{defi} A relation \(\leq\) on open automata is
\defitem \textbf{reflexive} iff \(\forall a, a \leq a\);
\defitem \textbf{transitive} iff \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\defitem \textbf{a preorder} iff it is reflexive and transitive;
\defitem \textbf{correct w.r.t.\@ composition} iff \(\forall a\, b, a\subst{b}{j} \leq a\);
\defitem \textbf{complete w.r.t.\@ composition} iff \(\forall a\, b, a \leq b \implies \exists c, a \cong b\subst{c}{j}\);
\defitem \textbf{context refining for composition} iff \(\forall a\, b\, c, a \leq b \implies a\subst{c}{j} \leq b\subst{c}{j}\);
\defitem \textbf{congruent for composition} iff \(\forall a\, b\, c, a \leq b \implies c\subst{a}{j} \leq c\subst{b}{j}\);
\defitem \textbf{compatible with composition} iff \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\subst{a}{j} \leq d\subst{b}{j}\);
\defitem \textbf{compatible with FH-bisimulation} iff \(\forall a\, b\, c, d, a \cong b \wedge c \cong d \wedge a \leq c \implies b \leq d\).
\end{defi}
Reflexivity, transitivity and preorder are classical properties on relations.
Correctness w.r.t.\@ composition means that every composition is considered a refinement.
This property captures the expected behaviour of a refinement relation on a compositional structure like the open automata.
Completeness w.r.t.\@ composition means that a refinement corresponds to the left automaton being equivalent to some composition of the right automaton.
This property is useful if we want to characterise the kind of operations that can generate any refined automaton starting from the specification.
We won't have it because extending the domain of definition of the initial valuation will be a refinement which can't be expressed with composition.
Yet a ``completeness w.r.t.\@ some set of operations" is still a strong and interesting property.
Context refinement is the refinement being compatible with composition of the same automaton.
Congruence is the refinement being compatible with being composed in the same automaton.
Compatibility with composition is the conjunction of the two latter (assuming the relation is a preorder).
This property is required on compositional structures to be of any use in verification, proofs and for model checking.
Finally compatibility with FH-bisimulation is the refinement not being able to distinguish two FH-bisimilar automata.
Since FH-bisimulation is the pre-existing equivalence relation, we should not introduce an incompatible proof strategy.

\subsection{Properties of the relation on states}
The important properties we consider for the relation on configurations are being a pre-simulation and preventing the introduction of deadlocks.
They are defined in the following.
\begin{defi}[Pre-simulation]
Let \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\).
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
This definition translated to open automata the classical definition of simulation \cite{10.1007/3-540-54430-5_78}.
The last formula means that for every pair of related configurations of the automaton and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update. % OR variableS
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we will assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between the related automata's holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n}\) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition that we put on any relation that we call a ``simulation".
